### 2021-08-25 plos_latex_template.tex
RJ: Added who the paper is for in the intro

In July 2021, the authors of this article, along with a larger group of people, organized the annual user conference of the R Project for Statistical Computing. This instance of the useR! conference was particularly well received as a virtual and explicitly global event, reaching users and developers of the R language from more than 120 countries. useR! conferences are non-profit events, organized by volunteers from the R community and supported by the R Foundation. The conference attracts a broad range of participants from academia, industry, government and the non-profit sector. For 2021, we aimed to build a high-quality conference in a kind, inclusive, accessible, and welcoming environment for everyone. In this article, we share a few critical lessons learned in the process. We streamline our most important learnings in 10 simple rules to host an inclusive conference. These rules apply equally to academic, industry, or mixed conferences; the rules are inspired by a global experience but also apply at the regional or local level.

@@

Calls to improve diversity in many spaces, including conferences, are becoming more frequent as communities and scientific societies are getting called out for over-representing specific groups. While there has been some action to address systematic inequalities, there is still a lot of room for improvement.

This article suggests rules to pivot traditional conferences towards inclusiveness and diversity and to welcome minoritized groups. These tips stem from the authors' experience of organizing useR! 2021, a virtual and global conference for users and developers of the R programming language \cite{r_core_team_2021}. We embraced the challenge of organizing a high-quality conference in the context of the COVID-19 pandemic using the chance of a virtual conference environment to create a kind, truly inclusive, accessible, and welcoming experience for everyone. Most of the authors also have experience of organizing other regional and national academic conferences and events in communities of practice such as R-Ladies and The Carpentries. Here, we share the lessons learned within the past year of organizing this global useR! conference, summarized as 10 simple rules towards an inclusive conference.

\section{Rule X: Define what diversity and inclusion mean for your conference}
%
The first step towards a diverse and inclusive conference is to recognize that people are diverse, that some groups face discrimination and might be underrepresented in your community and event \cite{timperley_he_2020}.

@@

Some teams gather this vision in a diversity statement (e.g., \url{https://user2021.r-project.org/about/diversity_statement/}). %and expressing this welcoming spirit in your communication strategy (social networks, website, brand, and visual identity) and in every decision you make, will let people know that they are seen, respected, and welcome; that this is their space and community too. 
However, it is important to not stop with a statement alone, but to identify specific goals and tasks that ensure compliance with these larger goals, for example, identifying accessible platforms, developing a detailed code of conduct, and preparing accessibility guidelines (see other Rules below).

%Lack of representation, unwelcoming -or overtly aggressive- environments hinder participation (or future participation) of people who could otherwise become active community members. In extreme cases, they can divert career paths, affect lives, and exclude people from some fields. So take this seriously, think about the reasons why you want to create a more diverse conference, and be open to learn. %Do not treat diversity as a checklist, do not use people from minoritized groups for image purposes (tokenization), and do not act as a savior but as a genuine collaborator. %ast: some version of this might come in a box.

\section{Rule X: Have a strong online component of the conference}

In-person interaction is priceless; however, it is more expensive for some, even unattainable for others. This inaccessibility is particularly true for global conferences that usually take place in high-income countries, making it financially demanding for international participants and often impossible to attend due to immigration requirements \cite{arend_disparity_2019, gewin_what_2019}. Online conferences are more inclusive: they do not need a visa or a big budget, and are more accessible to people who may be unable to travel because of health issues or family responsibilities. This means that online conferences have a greater reach, not only in terms of participants but in terms of the tutors and presenters that can participate \cite{atkinson_r_medicine_2021}. Furthermore, an online format is more environmentally friendly since it eliminates travel-related emissions \cite{sarabipour_evaluating_2020, niner_better_2021}; in a study case, conference attendance accounted for $35\%$ of a researchers footprint .

@@

\section{Rule X: Have an inclusive and diverse organizing team}
\label{rule_organizing_team}
A genuinely inclusive conference can only be organized by an inclusive and diverse organizing team. Build a team with people from different regions, genders, socioeconomic statuses, and other aspects of diversity. Particular attention should be paid to the usually underrepresented groups (see \textbf{Rule 1} discussing the dimensions of diversity lacking in your community and event). To ensure a deeper understanding of the challenges in different diverse groups, it is essential to create a representative working group that functions as a snapshot of the community at large. But this will only work if there is real inclusion. One would be surprised by how often conference leads tend to do everything themselves instead of reaching out to the right people for the right task -- likely because the latter involves careful planning and communication well ahead of time or because they naively overestimate their expertise in areas they lack practical experience in. For example, a cisgender or abled person estimating what might work best for transgender people or people with sensory disabilities. The same holds for other dimensions of diversity. A truly inclusive and welcoming space is one in which everyone in the team is invited and allowed to bring their experience to bear. Even if creating and maintaining such a team and space is challenging, the positive outcomes far outweigh any minor inconveniences (e.g. time zones), and the team will grow, learn, and embrace more inclusive practices \cite{hong_groups_2004}.

@@

%%Disabled people often repeat "Nothing about us without us", and this is true for every marginalized group. This means the actual life experiences, expertise, and insights from people in marginalized groups are not replaceable with good intentions from people outside these groups. No space will be truly inclusive and welcoming without people from these groups bringing their experience to bear. This means people in privileged groups may need to step down from positions of power. %this might
%rj: Yani's refs to have just in case:
%https://sites.lsa.umich.edu/scottepage/wp-content/uploads/sites/344/2015/11/pnas.pdf; https://www.pnas.org/content/early/2014/11/13/1407301111; https://www.mckinsey.com/business-functions/organization/our-insights/why-diversity-matters

\section{Rule X: Make the conference accessible to all, from start to end}

Conferences are often inaccessible to people with hearing, visual, and other disabilities \cite{price_access_2009}. %People may also find conferences inaccessible due to their caring responsibilities, dietary restrictions, religious practices, or LGBTQ+ status.
Inaccessible practices exclude people from conferences, while accessible practices are inclusive and can be beneficial to a broad spectrum of people. For instance, having captions is helpful to deaf and hard-of-hearing people, non-native speakers, and everyone in general.

If the conference is in person, the venue must be accessible for people with mobility limitations. 
Accommodations such as a quiet space for neurodivergent people and a space for breastfeeding should also be provided.
Presenters should always speak into a microphone to make it easier for the hard of hearing and for captioners or interpreters to listen to them. 
Regardless of the conference format --online, in person, or hybrid --images used in communications about the conference should be accompanied by alternative text, while videos should have both captions and sound. 
Platforms for conference registration and abstract submission, websites, and chat platforms --if used --should be screen-reader friendly and keyboard accessible, with low technology requirements (hardware, software, and internet connection). 
Captioning for presentations should be available in more than one language if possible.
These aspects should be tested well in advance of going live.

@@

\section{Rule X: Have a code of conduct and a team to reinforce it}

@@

\section{Rule X: Search proactively the best people for invited roles}

When inviting people as keynote speakers, program committee, session chairs and others, we should always look for the best in their subject. Ensuring diversity in each of these roles is a deliberate process -- if there is not much diversity in the first set of names, it needs to be revised to include people from minoritized groups. If these groups are not showing up in the A-list, it is seldom because of their technical acumen -- it is because we, as organizers, are not used to looking in unfamiliar places to bring in the talent. Many of our biased lists are products of the existing systems that have always privileged some groups of people \cite{dwyer_notice_2021, sarabipour_evaluating_2020}. Rather than deter us, this should encourage us to look further to find great people that are not routinely in the spotlight. We need to go beyond our narrow and often limited networks to look for, reach out to, invite, encourage, and onboard these people until there is ample representation across the diversity spectrum and dimensions.

% So, as long as we start with an inclusive and diverse organizing team (\textbf{Rule 2}), we can aim at finding the right people for various aspects of the conference organization. The regional and local communities are also good sources to tap into. For example, for useR! 2021, groups like AfricaR (africa-r.org), R-Ladies (rladies.org), MiR (mircommunity.com), Forwards (forwards.github.io), and LatinR (latin-r.com) were fundamental to reach people for the organizing team, potential attendees, and sponsors.

@@

%Representation is an important aspect of diversity and inclusivity: seeing people like you, from your country, that speak your language is one of the best way to feel you belong, this is why the rates can not only be different for registration, but also, for sponsors, this allows us to have institutions and companies from all over the world being part of the conference which also means they will communicate about our event in their local communities.

%rationale: 1. they might select for the already privileged or people who dont need those to be nice on the cvs 2. they select for outgoing extroverted privileged ("brave") people 3. they are time and emotionally consuming which selects out some people 4. they single out people. even if you controlled for 1 to 3 you would "badge" people - you are creating a minority while trying to avoid it.
Offering scholarships or grants to attend the conference may be counterproductive, as they can be seen as more of a competition and attract people that might not need the money (e.g., when their PI could pay for it) but could apply for it to enrich their CV while filtering out people who do not feel entitled to earn them. If the goal is to help the people who really need it, the language needs to make that clear without shrouding it with fancy-sounding phrases or tokenizing phrases (e.g. `grant recipient' or `diversity scholar', respectively). Do not call it a scholarship or a grant. Ultimately, it is a fee waiver or a discount for the people who need additional financial support. Also, the process for applicants should be simplified. Applying for loans, grants, and scholarships is already an emotionally demanding task. Do not complicate the attendees' lives by asking for long essays to convince a committee that they deserve your support. For this stage of the process, a certain degree of trust goes a long way -- when they say they need the support, it is best to give them the waiver rather than second guess their eligibility. \st{Also, the participant metadata (e.g., demographics, gender, career stage, disabilities, if they disclose them) is already a strong indicator of their potential need for aid.} Other If the conference resources allow for it, you could even take further steps to offer financial support for activities that help them have the time and resources to be at the conference: child care support, transportation fees (if in-person), or internet connection services (if virtual).

@@

\section{Rule X: Take deliberate efforts to promote the conference among underrepresented groups}

If part of the community has been historically discriminated against, one should emphasize that they are particularly welcome in this event and that the organization will make it a safe and inclusive environment. Publish a diversity statement (e.g., \url{https://user2021.r-project.org/about/diversity_statement/}), the code of conduct and accessibility guidelines. Fee waivers, financial support, and ease of application/registration should also be advertised; people who are used to applying and getting rejected for scholarships and waivers may find it relieving that the process will be supportive rather than discriminatory against underrepresented groups. Advertise the conference in multiple languages (rule \ref{rule_language}). Express this welcoming spirit in your communication strategy (social networks, website, brand, and visual identity) to let people know that they are seen, respected, and welcome; that this is their space and community too.

%Providing financial and other support (e.g., childcare, flexible hours) will help minoritized groups partake more willingly in the organization and the actual conference.
%ast likewise, some key people in your organizing team might not be able to collaborate without remuneration. Resources permitting, it may be a good idea to secure funds to offer stipends to them.

@@

@@

The ten rules stated here can be adapted depending on the conference format and settings. From our own experience, we are aware that enforcing these rules requires a tremendous amount of work. We engaged in most of these practices, and learned others too late, like having an explicit payment category for postdocs or paying the whole organizing team, so we share them here as part of our learning process. 
Since accessibility and inclusiveness are not as widespread as they should be, the process might seem exacting. But this is where the power of communities and individuals reside. If more conferences and domains employ these techniques, the process will get more streamlined, straightforward, and mainstream to adapt with minimal overhead. And we will make a change towards healthier, stronger, and more inclusive communities, empowering historically marginalized people and learning from them, thereby increasing the overall quality of our conferences.%After all, the strength of any community lies in its ability to be adaptive, inclusive, and accessible.

>>>>

In July 2021, the authors of this article participated in the organization team of the annual user conference of the R Project for Statistical Computing. useR! conferences are non-profit events, organized by volunteers from the R community and arranged by the R Foundation. The conference attracts a broad range of participants from academia, industry, government, and the non-profit sector. For 2021, we aimed to build a high-quality virtual and explicitly global conference in a kind, inclusive, accessible, and welcoming environment for everyone. 
In this article, we share a few lessons learned in the process. We streamline our most important learnings in 10 simple rules to host an inclusive conference. These rules apply equally to academic, industry, or mixed conferences; the rules are inspired by a global experience but also apply at the regional or local level.

%ast: matt's original sentence: reaching users and developers of the R language from more than 120 countries.

@@

%ast will write a paragraph about why diversity and the systemic inequalities
While there has been some action to address systematic inequalities, there is still a lot of room for improvement.

This article suggests rules to pivot traditional conferences towards inclusiveness and diversity and to welcome minoritized groups. It is directed as people who are part of a stable meetings commitee that oversees the site/location selection process or that coordinates with the local organizers (e.g. R Foundation Conference Committee, Ecological Society of America Meetings Committee). It is also directed at the local/virtual organizers who desire to make an inclusive conference starting at the planning stage.

These tips stem from the authors' experience of organizing useR! 2021, a virtual and global conference for users and developers of the R programming language \cite{r_core_team_2021}. We embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, truly inclusive, accessible, and welcoming experience for everyone. Most of the authors also have experience of organizing other regional and national academic conferences and events in communities of practice such as R-Ladies and The Carpentries. Here, we share the lessons learned within the past year of organizing this global useR! conference, summarized as 10 simple rules towards an inclusive conference.

\section{Rule X: Define what diversity and inclusion mean for your conference}
\label{rule_diversity}

The first step towards a diverse and inclusive conference is to recognize that people are diverse, that some groups face discrimination and might be underrepresented in your community and event \cite{timperley_he_2020}.

@@

To accomplish this diversity vision, you will have to take some conscious decisions toward removing bias from your conference's spaces. Having diverse people in decision-making positions will affect positively all the other aspects of your conference, not only internally because all the decision-making process will benefit from their input \ref{rule_organizing_team}, but also externally because representation matters \ref{rule_communication}.

While you may want to improve representation towards some of the most visible dimensions of human diversity, such as race, gender, and country of origin, building a truly inclusive environment means taking care of all the other aspects as well. Having consideration of religious practices, setting specific accommodations for lactating women and child care, having LGBTIQ+-friendly spaces, community-only spaces, enforcing the use of pronouns, and treating gender as a non-binary variable, are some examples of decisions that can make inclusion real. 
%Representation is an important aspect of diversity and inclusion: seeing people like you, from your country, that speak your language is one of the best way to feel you belong
Some teams gather this vision in a diversity statement (e.g., \url{https://user2021.r-project.org/about/diversity_statement/}). %and expressing this welcoming spirit in your communication strategy (social networks, website, brand, and visual identity) and in every decision you make, will let people know that they are seen, respected, and welcome; that this is their space and community too. 
However, it is important to not stop with a statement alone, but to identify specific goals and tasks that ensure compliance with these larger goals, for example, identifying accessible platforms, developing a detailed code of conduct and a team to enforce it, and preparing accessibility guidelines (see other Rules below).

Lack of representation, unwelcoming -or overtly aggressive- environments hinder participation (or future participation) of people who could otherwise become active community members. In extreme cases, they can divert career paths, affect lives, and exclude people from some fields. So take this seriously, think about the reasons why you want to create a more diverse conference, and be open to learn. Do not treat diversity as a checklist, do not use people from minoritized groups for image purposes (tokenization), and do not act as a savior but as a genuine collaborator. %ast: some version of this might come in a box.

\section{Rule X: Have a strong online component of the conference} 
\label{rule_online}
In-person interaction is priceless; however, it is more expensive for some, even unattainable for others. This inaccessibility is particularly true for global conferences that usually take place in high-income countries, making it financially demanding for international participants and often impossible to attend due to immigration requirements \cite{arend_disparity_2019, gewin_what_2019}. Online conferences are more inclusive: they do not need a visa or a big budget, and are more accessible to people who may be unable to travel because of health issues or family responsibilities. This means that online conferences have a greater reach, not only in terms of participants but in terms of the tutors and presenters that can participate \cite{atkinson_r_medicine_2021}. Furthermore, an online format is more environmentally friendly since it eliminates travel-related emissions \cite{sarabipour_evaluating_2020, niner_better_2021}; in a study case, conference attendance accounted for $35\%$ of a researcher's footprint \cite{achten_carbon_2013}.

@@

\section{Rule X: Have an inclusive and diverse organizing team}
\label{rule_organizing_team}
A genuinely inclusive conference can only be organized by an inclusive and diverse organizing team. Build a team with people from different regions, genders, socioeconomic statuses, and other aspects of diversity. Particular attention should be paid to the usually marginalized groups (see \textbf{Rule 1} discussing the dimensions of diversity lacking in your community and event). To ensure a deeper understanding of the challenges in different diverse groups, it is essential to create a representative working group that functions as a snapshot of the community at large. But this will only work if there is real inclusion. One would be surprised by how often conference leads tend to do everything themselves instead of reaching out to the right people for the right task -- likely because the latter involves careful planning and communication well ahead of time or because they naively overestimate their expertise in areas they lack practical experience in. For example, a cisgender or abled person estimating what might work best for transgender people or people with sensory disabilities. The same holds for other dimensions of diversity. A truly inclusive and welcoming space is one in which everyone in the team is invited and allowed to bring their experience to bear. Even if creating and maintaining such a team and space is challenging, the positive outcomes far outweigh any minor inconveniences (e.g. time zones), and the team will grow, learn, and embrace more inclusive practices \cite{hong_groups_2004}.

@@

%%Disabled people often repeat "Nothing about us without us", and this is true for every marginalized group. This means the actual life experiences, expertise, and insights from people in marginalized groups are not replaceable with good intentions from people outside these groups. 
%rj: Yani's refs to have just in case:
%https://sites.lsa.umich.edu/scottepage/wp-content/uploads/sites/344/2015/11/pnas.pdf; https://www.pnas.org/content/early/2014/11/13/1407301111; https://www.mckinsey.com/business-functions/organization/our-insights/why-diversity-matters

\section{Rule X: Make the conference accessible to all, from start to end}
\label{rule_accessibility}
Conferences are often inaccessible to people with hearing, visual, and other disabilities \cite{price_access_2009}. %People may also find conferences inaccessible due to their caring responsibilities, dietary restrictions, religious practices, or LGBTQ+ status.
Inaccessible practices exclude people from conferences, while accessible practices are inclusive and can be beneficial to a broad spectrum of people. For instance, having captions is helpful to deaf and hard-of-hearing people, non-native speakers, and everyone in general.

If the conference is in person, the venue must be accessible for people with mobility limitations. 
Accommodations such as a quiet space for neurodivergent people should also be provided.
Presenters should always speak into a microphone to make it easier for the hard of hearing and for captioners or interpreters to listen to them. 
Regardless of the conference format --online, in person, or hybrid --images used in communications about the conference should be accompanied by alternative text, while videos should have both captions and sound. 
Platforms for conference registration and abstract submission, websites, and chat platforms --if used --should be screen-reader friendly and keyboard accessible, with low technology requirements (hardware, software, and internet connection). 
Captioning for presentations should be available in more than one language if possible.
These aspects should be tested well in advance of going live.

@@

\section{Rule X: Adopt a code of conduct and create a team to reinforce it}
\label{rule_CoC}

@@

\section{Rule X: Search proactively for the best people the conference's vision}
\label{rule_presenters}
When inviting people as keynote speakers, program committee, session chairs and others, we should always look for the best in their subject. Ensuring diversity in each of these roles is a deliberate process -- if there is not much diversity in the first set of names, it needs to be revised to include people from minoritized groups. If these groups are not showing up in the A-list, it is seldom because of their technical acumen -- it is because we, as organizers, are not used to looking in unfamiliar places to bring in the talent. Many of our biased lists are products of the existing systems that have always privileged some groups of people \cite{dwyer_notice_2021, sarabipour_evaluating_2020}. Rather than deter us, this should encourage us to look further to find great people that are not routinely in the spotlight. We need to go beyond our narrow and often limited networks to look for, reach out to, invite, encourage, and onboard these people until there is ample representation across the diversity spectrum and dimensions.

% So, as long as we start with an inclusive and diverse organizing team (\textbf{Rule 2}), we can aim at finding the right people for *these invited roles*. The regional and local communities are also good sources to tap into. For example, for useR! 2021, groups like AfricaR (africa-r.org), R-Ladies (rladies.org), MiR (mircommunity.com), Forwards (forwards.github.io), and LatinR (latin-r.com) were fundamental to reach people for the organizing team, potential attendees, and sponsors. %ast: i think this could come back, removing the specific part about "conference organization" - i edited that.

@@

%rationale: 1. they might select for the already privileged or people who dont need those to be nice on the cvs 2. they select for outgoing extroverted privileged ("brave") people 3. they are time and emotionally consuming which selects out some people 4. they single out people. even if you controlled for 1 to 3 you would "badge" people - you are creating a minority while trying to avoid it.
Offering scholarships or grants to attend the conference may be counterproductive, as they can be seen as more of a competition and attract people that might not need the money (e.g., when their PI could pay for it) but could apply for it to enrich their CV while filtering out people who do not feel entitled to earn them. If the goal is to help the people who really need it, the language needs to make that clear without shrouding it with fancy-sounding phrases or terms that single out the recipients (e.g. `grant recipient' or `diversity scholar', respectively). Do not call it a scholarship or a grant. Ultimately, it is a fee waiver or a discount for the people who need additional financial support. Also, the process for applicants should be simplified. Applying for loans, grants, and scholarships may be an emotionally demanding task. Do not complicate the attendees' lives by asking for long essays to convince a committee that they deserve your support. For this stage of the process, a certain degree of trust goes a long way -- when they say they need the support, it is best to give them the waiver rather than second guess their eligibility. If the conference resources allow for it, you could even take further steps to offer financial support for activities that help them have the time and resources to be at the conference: child care support, transportation fees (if in-person), or internet connection services (if virtual).

@@

\section{Rule X: Actively reach out and involve people who have been systematically excluded}
\label{rule_communication}
%this rule name should be more explicit towards communication, "promoting"
If part of the community has been historically discriminated against, one should emphasize that they are particularly welcome in this event and that the organization will make it a safe and inclusive environment. Publish a diversity statement (e.g., \url{https://user2021.r-project.org/about/diversity_statement/}), the code of conduct and accessibility guidelines. Fee waivers, financial support, and ease of application/registration should also be advertised; people who are used to applying and getting rejected for scholarships and waivers may find it relieving that the process will be supportive rather than discriminatory against underrepresented groups. Advertise the conference in multiple languages (rule \ref{rule_language}). Express this welcoming spirit in your communication strategy (social networks, website, brand, and visual identity) to let people know that they are seen, respected, and welcome; that this is their space and community too.

@@

\section{Rule X: Don't let perfect be the enemy of good}
 While conference teams should have a bold vision and clear targets consensus about where to shine and where to compromise is equally important. As a team, this is a challenging journey. It means letting go of some ideals, or committing time and money into something that one does not find particularly useful or important.

Remember, aspirations can be local and narrow and as a consequence shy away valuable contributors. A late bus is almost certainly better than no bus. A broader pool of contributors does not only allow organizers to add more and work on multiple aspects at a different pace, it also helps to split workload and responsibilities to avoid burning out over the marathon that organizing a conference is.

Without allowing for compromises, workload will naturally channel to those who are committed the most and already carry the largest share of the load. Hence compromises are no sign of giving in but are crucial to delegation, building self-responsible teams and keep a good spirit in the long run.

@@

The ten rules stated here can be adapted depending on the conference format and settings. From our own experience, %it's worth it :P
we are aware that enforcing these rules requires a tremendous amount of work. We engaged in most of these practices, and learned others too late, so we share them here as part of our learning process. 
Since accessibility and inclusiveness are not as widespread as they should be, the process might seem exacting. But this is where the power of communities and individuals reside. If more conferences and domains employ these techniques, the process will get more streamlined, straightforward, and mainstream to adapt with minimal overhead. And we will make a change towards healthier, stronger, and more inclusive communities, empowering historically marginalized people and learning from them, thereby increasing the overall quality of our conferences.%After all, the strength of any community lies in its ability to be adaptive, inclusive, and accessible.
